---
title: 'spatial weights: Rook Contiguity Weights'
output: html_document
---

# TL;DR 

The rook rule defines neighbors by the existence of a common edge between 02 spatial units.

Spatial weights are central components in spatial data analysis. The spatial weights represent the possible spatial interactions between observations in space. rgeoda provides 6 functions to create 4 different types of spatial weights:

- Contiguity Based Weights
+ (1) Contiguity Based Weights: queen_weights(), rook_weights()

- Distance Based Weights
+ (1) Distance Based Weights: distance_weights()
+ (2) K-Nearest Neighbor Weights: knn_weights()
+ (3) Kernel Weights: distance_weights() and knn_weights() with kernel parameters

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load required packages 

```{r message=FALSE, warning=FALSE}
library(rgeoda)
library(sf)
```

# Load shapefile 

```{r}
guerry_path <- system.file("extdata", "Guerry.shp", package = "rgeoda")
guerry <- st_read(guerry_path) # spatial object
```

# Spatial Weights

## Spatial weight matrix (rook contiguity rule)

```{r}
rook_w <- rook_weights(guerry)
summary(rook_w)
```

```{r}
# Assess attributes of weights obj
is_symmetric(rook_w)
has_isolates(rook_w) 
weights_sparsity(rook_w) # ideal ~ 1 (0,6 - 0,8)
```

## Calculation examples 

```{r}
# To access the details of the weights: e.g. list the neighbors of a specified observation:
nbrs <- get_neighbors(rook_w, idx = 1)
cat("\nNeighbors of the 1-st observation are:", nbrs)
```

```{r}
# To compute the spatial lag of a specified observation by passing the values of the selected variable
# This function calculates the spatial lag of the variable Crm_prs (which represents crimes against persons).
lag <- spatial_lag(rook_w, guerry['Crm_prs'])
lag # numeric vector where each entry represents the spatially lagged crime rate for each département
```

## Mapping 

```{r}
# R workflow 
library(sf)       # For spatial data handling
library(spdep)    # For spatial weights and spatial lag computation
library(tmap)     # For visualization

# Load dataset if available
data(guerry, package = "spData")  

# Ensure guerry is an sf object
guerry_sf <- st_as_sf(guerry)

# Compute spatial lag of 'Crm_prs'
guerry_sf$Crm_prs_lag <- spatial_lag(rook_w, guerry_sf['Crm_prs'])

# Create a choropleth map to visualize the spatial lag
r_map <- tm_shape(guerry_sf) +
  tm_polygons("Crm_prs_lag", 
              title = "Spatial Lag of Crime Against Persons",
              palette = "Red",  # Use a color gradient
              style = "quantile") + 
  tm_layout(legend.outside = TRUE)

r_map 
# save
tmap_save(r_map,
          filename = "./plot/spatial_weights/r_weights_1.png", 
          width = 8, height = 6, dpi = 300)
```

A map of France where different shades of red represent varying levels of crime spillover from neighbouring regions.

```{r}
# Revise the map 
# Define a color scheme (choose one)
color_schemes <- c("Blues", "Reds", "Viridis", "PuBuGn", "YlOrBr")
chosen_palette <- "Viridis"  # Change this to any scheme from above

# Create a choropleth map with labels
r_map <- tm_shape(guerry_sf) +
  tm_polygons("Crm_prs_lag", 
              title = "Spatial Lag of Crime Against Persons",
              palette = chosen_palette,  # Apply the chosen color scheme
              style = "quantile", 
              border.col = "black",  # Add borders for clarity
              lwd = 0.5) +  # Adjust border thickness
  tm_text("Dprtmnt", size = 0.7, col = "black") +  # Add region labels
  tm_layout(legend.outside = TRUE,
            main.title = "Crime Spillover Effect in Neighbouring Regions",
            main.title.size = 1.2,
            frame = FALSE)  # Remove outer frame for a clean look
r_map

# save
tmap_save(r_map,
          filename = "./plot/spatial_weights/r_weights_2.png", 
          width = 8, height = 6, dpi = 300)
```

## Save the weight matrix

```{r}
# save weights
save_weights(rook_w, 
             guerry['CODE_DE'], 
             out_path = './output/Guerry_rook_w.gal', 
             layer_name = 'Guerry')
```

