---
title: 'spatial weights: Distance Based Weights'
output: html_document
---

# TL;DR 

Spatial weights are central components in spatial data analysis. The spatial weights represent the possible spatial interactions between observations in space. rgeoda provides 6 functions to create 4 different types of spatial weights:

- Contiguity Based Weights
+ (1) Contiguity Based Weights: queen_weights(), rook_weights()

- Distance Based Weights
+ (1) Distance Based Weights: distance_weights()
+ (2) K-Nearest Neighbor Weights: knn_weights()
+ (3) Kernel Weights: distance_weights() and knn_weights() with kernel parameters

# Library 

```{r message=FALSE, warning=FALSE}
library(rgeoda)
library(sf)
```

# Load shapefile 

```{r}
guerry_path <- system.file("extdata", "Guerry.shp", package = "rgeoda")
guerry <- st_read(guerry_path) # spatial object
```

# Spatial Weights

## Distance Based Weights

The most straightforward spatial weights matrix constructed from a distance measure is obtained when i and j are considered neighbors whenever j falls within a critical distance band from i. 

Distance-Based Weights assigns weights based on a fixed distance threshold, where regions within the specified distance are considered neighbors, often used for capturing local spatial interactions, or being used when spatial influence is assumed to have a fixed cutoff, such as environmental monitoring where pollutants affect areas within a specific radius.

To start the distance based neighbors, a threshold value must be computed. 'rgeoda' provides a function min_distthreshold helping to find a optimized distance threshold that guarantees that every observation has at least one neighbor. 

```{r}
dist_thres <- min_distthreshold(guerry)
dist_thres
```

```{r}
dist_w <- distance_weights(guerry, dist_thres)
summary(dist_w)
```

## K-Nearest Neighbor Weights

The rule is very obersvation will have exactly k neighbors. It can be used to avoid the problem of isolate in distance-band weights when a smaller cut-off distance is used. 

The method is useful for handling uneven spatial distributions, datasets with varying spatial densities, such as economic networks or social interactions.

```{r}
# For example, to create a 6-nearest neighbor weights using Guerry:
knn6_w <- knn_weights(guerry, 6)
summary(knn6_w)
```

## Kernel Weights

Kernel weights apply kernel function to determine the distance decay (decreasing of distance) in the derived continuous weights kernel. 

This method applies a kernel function to model distance decay, allowing for a smooth, continuous weighting of spatial relationships rather than binary adjacency. Best for modeling gradual distance decay effects, often used in epidemiology, transportation modeling, and diffusion studies.

Kernel function include:
- Triangular (hình tam giác)
- Uniform (đồng nhất)
- Quadratic (bậc hai)
- Epanechnikov
- Quartic (bậc bốn)
- Gaussian (chuẩn)

```{r}
# To create a kernel weights with fixed bandwith:
bandwidth <- min_distthreshold(guerry)

kernel_w <- kernel_weights(guerry, 
                           bandwidth, 
                           kernel_method = "uniform")
summary(kernel_w)
```

```{r}
# To create a kernel weights with adaptive bandwidth or using max Knn distance as bandwidth:
adptkernel_w = kernel_knn_weights(guerry, 
                                  6, 
                                  "uniform")
summary(adptkernel_w)
```

