---
title: "spatio-temporal fitting model"
output: html_document
---

# Dependencies 

```{r}
library(tidyverse)
library(kableExtra)
library(sf) 
library(sp) # spatial object conversion 
library(tmap) 
library(viridis) 
library(corrplot)
library(gghighlight) # highlight data on plot 

# Analysing spatio-temporal data
library(spacetime)
library(lubridate)
library(MASS) # applied statistics 
library(lmtest) # statistical test for linear regression model 
library(FRK) # fit spatial random effect models 
library(jtools) # Exportable regression tables
```

# Prepare data

```{r}
# remove geometries
st_geometry(covid19_spt) <- NULL

# share of population in long-term illness 
covid19_spt <- covid19_spt %>% 
  mutate(lt_illness = Longterm_sick_or_disabled / Residents
)
```

# Construct Basic Function

```{r}
# build basis functions
G <- auto_basis(data = covid19_spt[,c("long","lat")] %>%
                       SpatialPoints(),           # To sp obj
                nres = 1,                         # One resolution
                type = "Gaussian")                # Gaussian BFs
```

```{r}
# basis functions evaluated at data locations are then the covariates
S <- eval_basis(basis = G,                       
                s = covid19_spt[,c("long","lat")] %>%
                     as.matrix()) %>%            
     as.matrix()                                 
```

```{r}
colnames(S) <- paste0("B", 1:ncol(S)) # assign column names
```

# Add basic function to dataframe

```{r}
# Prepare a data frame for the regression model
# adding the weights extracted from the basis functions. These weights enter as covariates in model
# selecting variables
reg_df <- cbind(covid19_spt, S) %>%
  dplyr::select(ctyu19nm, n_covid19_r, long, lat, date, lt_illness, B1:B9)
```

# Fit linear regression 

```{r}
eq1 <- n_covid19_r ~ long + lat + date + lt_illness + .

lm_m <- lm(formula = eq1, 
           data = dplyr::select(reg_df, -ctyu19nm))

lm_m %>% summary()
```

# Poisson Regression

```{r}
# A key issue: dependent variable is a count and is highly skewed
# address challenges by using poisson regression 

# estimate a poisson model
poisson_m1 <- glm(eq1,
                family = poisson("log"), # Poisson + log link
                data = dplyr::select(reg_df, -ctyu19nm))
poisson_m1 %>% summary()
```

```{r}
# Challenges: overdispersion (ratio > 1)
poisson_m1$deviance / poisson_m1$df.residual
```

# Quasipoisson Regression

```{r}
# estimate a quasipoisson model to deal with over-dispersion 
qpoisson_m1 <- glm(eq1,
                family = quasipoisson("log"), 
                data = dplyr::select(reg_df, -ctyu19nm))
qpoisson_m1 %>% summary()
```

# Negative Binomial Regression

```{r}
# alternative solution for overdispersion 
# estimate a negative binomial model
nb_m1 <- glm.nb(eq1, 
       data = dplyr::select(reg_df, -ctyu19nm))
nb_m1 %>% summary()
```

# Negative Binomial Regression - Including Interactions

```{r}
# long + lat + date + long:lat + long:date + lat:date
# modelling how the impact of space changes over time, and vice versa

# new model specification
eq2 <- n_covid19_r ~ (long + lat + date)^2 + lt_illness + .
# estimate a negative binomial model
nb_m2 <- glm.nb(eq2, 
       data = dplyr::select(reg_df, -ctyu19nm))
nb_m2 %>% summary()
```

# Model comparision 

To get a quick sense of model fit is to explore the correlation between observed and predicted values of the dependent variable


```{r}
# computing predictions for all models
lm_cnt <- predict(lm_m)
poisson_cnt <- predict(poisson_m1)
nb1_cnt <- predict(nb_m1)
nb2_cnt <- predict(nb_m2)

reg_df <- cbind(reg_df, lm_cnt, poisson_cnt, nb1_cnt, nb2_cnt)
```

```{r}
# computing correlation coefficients
cormat <- cor(reg_df[, c("n_covid19_r", "lm_cnt", "poisson_cnt", "nb1_cnt", "nb2_cnt")], 
              use="complete.obs", 
              method="pearson")

# significance test
sig1 <- corrplot::cor.mtest(reg_df[, c("n_covid19_r", "lm_cnt", "poisson_cnt", "nb1_cnt", "nb2_cnt")],
                            conf.level = .95)

# create a correlogram
corrplot::corrplot.mixed(cormat,
                         number.cex = 1,
                         tl.pos = "d",
                         tl.cex = 0.9)
```


