---
title: "spatio-temporal - data preparation"
output: html_document
---


For this analyisi, we will use data on:
1. COVID-19 confirmed cases from 30th January, 2020 to 21st April, 2020 from Public Health England via the GOV.UK dashboard;
2. resident population characteristics from the 2011 census, available from the Office of National Statistics; and,
3. 2019 Index of Multiple Deprivation (IMD) data from GOV.UK and published by the Ministry of Housing, Communities & Local Government. The data are at the ONS Upper Tier Local Authority (UTLA) level - also known as Counties and Unitary Authorities.

# Loading Library 

```{r}
# List of required packages
packages <- c(
  "tidyverse", "kableExtra", "sf", "sp", "tmap", "viridis", 
  "corrplot", "gghighlight", "spacetime", "lubridate", "MASS", 
  "lmtest", "FRK", "jtools"
)

# Install any missing packages
installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed])
}

# Load packages
lapply(packages, library, character.only = TRUE)

```

# Data Wrangling 

## Data Preparation 

```{r}
# read ONS UTLA shapefile
utla_shp <- st_read("data/spatio-temporal/ons_utla.shp") 
```

```{r}
# create table of locations 
locs <- utla_shp %>% as.data.frame() %>% 
  dplyr::select(objct, cty19c, ctyu19nm, long, lat, st_rs)
```

```{r}
# read time data frame
time <- read_csv("data/spatio-temporal/reporting_dates.csv")
```

```{r}
# read COVID-19 data in long format
covid19 <- read_csv("data/spatio-temporal/covid19_cases.csv")
```

```{r}
# read 2011 census and 2019 Index of Multiple Deprivation (IMD)
censusimd <- read_csv("data/spatio-temporal/2011census_2019imd_utla.csv")
```

```{r}
head(covid19, 3)
```

```{r}
head(covid19$date, 5) #check time structure used for reporting covid cases 
```

```{r}
# parsing data into a time stamp
covid19$date <- ymd(covid19$date)
class(covid19$date)
```

```{r}
# separate date variable into day, week, month and year
covid19$day <- day(covid19$date)
covid19$week <- week(covid19$date) # week of the year
covid19$month <- month(covid19$date)
covid19$year <- year(covid19$date)
```

## Build a 'spatio-temporal object' of class STIDF using STIDF from the 'spacetime' package

```{r}
# add spatial information contained in shapefile to create a spatio-temporal data frame
covid19_spt <- left_join(utla_shp, 
                         covid19, 
                         by = c("ctyu19nm" = "Area.name"))
```

```{r}
class(covid19_spt)
```

```{r}
# identifying spatial fields
spat_part <- sp::as(
  dplyr::select(
    covid19_spt,
    -c(bng_e, bng_n, Area.code, Area.type, Daily.lab.confirmed.cases,
       Cumulative.lab.confirmed.cases, date, day, week, month, year)
  ),
  Class = "Spatial"
)

#'as(..., Class = "Spatial")' from 'sp' package was used to convert the selected sf or data.frame object into a Spatial object
```

```{r}
# identifying temporal fields
temp_part <- covid19_spt$date

# identifying data
covid19_data <- covid19_spt %>%
  dplyr::select(c(
    Area.code, Area.type, date, Daily.lab.confirmed.cases,
    Cumulative.lab.confirmed.cases, day, week, month, year
  )) %>%
  as.data.frame()
```

```{r}
# construct STIDF object
covid19_stobj <- STIDF(sp = spat_part, # spatial fields
                time = temp_part, # time fields
                data = covid19_data) # data
                
class(covid19_stobj)
```

```{r}
# Next add census and IMD variables

# select pop data
pop <- censusimd %>% dplyr::select("UTLA19NM", "Residents", "Longterm_sick_or_disabled") #For the purposes of this analysis, only add total population and long-term sick or disabled population counts

# join dfs
covid19_spt <- left_join(covid19_spt, pop,
                         by = c("ctyu19nm" = "UTLA19NM"))
covid19 <- left_join(covid19, pop, by = c("Area.name" = "UTLA19NM"))
```
