---
title: "spatio-temporal - exploratory analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mean

## Empirical Spatial Mean

```{r}
# compute empirical spatial mean
sp_av <- covid19_spt %>% group_by(ctyu19nm) %>% # group by spatial unit
  summarise(sp_mu_emp = mean(n_covid19_r))

# plot empirical spatial mean
ggplot(data=sp_av) +
  geom_col( aes( y = reorder(ctyu19nm, sp_mu_emp), x = sp_mu_emp) , fill = "grey50") +
  theme_classic() +
  labs(title= paste(" "), x="Average New Cases per 100,000", y="Upper Tier Authority Area") +
  theme(legend.position = "bottom") +
  theme(axis.text.y = element_text(size=7)) +
  theme(axis.text.x = element_text(size=12)) +
  theme(axis.title=element_text(size=20, face="plain"))
```

## Empirical Temporal Mean

```{r}
# compute temporal mean
tm_av <- covid19 %>% group_by(date) %>%
  summarise(tm_mu_emp = mean(n_covid19_r))

# plot temporal mean + trends for all spatial units
ggplot() +
  geom_line(data = covid19, mapping = aes(x =date, y = n_covid19_r,
                          group = Area.name), color = "gray80") +
   theme_classic() +
  geom_smooth(data = tm_av, mapping = aes(x =date, y = tm_mu_emp), 
              alpha = 0.5,
              se = FALSE) +
    labs(title= paste(" "), x="Date", y="Cumulative Cases per 100,000") +
    theme_classic() +
    theme(plot.title=element_text(size = 18)) +
    theme(axis.text=element_text(size=14)) +
    theme(axis.title.y = element_text(size = 16)) +
    theme(axis.title.x = element_text(size = 16)) +
    theme(plot.subtitle=element_text(size = 16)) +
    theme(axis.title=element_text(size=18, face="plain"))
```

# Dependence 

## Spatial dependence 
- Compute the Moran’s I on the rate of new COVID-19 cases (i.e. n_covid19_r in the covid19 data frame) at multiple time points

## Temporal dependence

Before measuring the temporal dependence is time-series, a time-series object needs to be created with a time stamp and given cycle frequency. 
- A cycle frequency refers to when a seasonal pattern is repeated. 
- We consider a time series of the total number of new COVID-19 cases per 100,000 (i.e. we sum cases over UTLAs by day) and the frequency set to 7 to reflect weekly cycles. So we end up with a data frame of length 71.

```{r}
# create a time series object
total_cnt <- covid19 %>% 
  group_by(date) %>%
  summarise(new_cases = sum(n_covid19_r)) 

total_cases_ts <- ts(total_cnt$new_cases,  
                     start = 1,
                     frequency =7)
# ts() is R Base to creates a regular time series object from a numeric vector
```

```{r}
# test temporal autocorrelation
# compute the correlation coefficient between a time series measured at time t and its lag measured at time t-1
# create lag term t-1
lag_new_cases <- total_cnt$new_cases[-1]
total_cnt <- cbind(total_cnt[1:70,], lag_new_cases)
cor(total_cnt[,2:3])
```

## Time series component

```{r}
# decompose time series
dec_ts <- decompose(total_cases_ts)
# plot time series components
plot(dec_ts)
```

