---
title: "Geographically Weighted Regression (GWR) Model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Geographically Weighted Regression

Geographically Weighted Regression (GWR) is a local regression model. It is the term introduced by Fotheringham, Charlton and Brunsdon (1997, 2002) to describe a family of regression models in which the coefficients are allowed to vary spatially. 

# Dataset  

```{r}
nycDat <- sf::st_read('./data/nyctract_acs/NYC_Tract_ACS2008_12.shp')
```

```{r}
sf::st_crs(nycDat) <- 4326
```

# GWR model 

## Bandwidth Calculation 

```{r}
nycDatNoNA <- nycDat %>% tidyr::drop_na()
```

```{r}
library(car)
library(GWmodel)

# Check multicollinearity
vif_vals <- vif(lm(log(popunemplo +1) ~ log(1+popinlabou) +
                      log(1+onlylessth) + log(1+master) + 
                      log(1+africaninl) + log(1+asianinlab) + 
                      log(1+hispanicin) + medianage, 
                    data = nycDatNoNA))

print(vif_vals) # threshold 10 -> remove 'medianage'
```

```{r}
# remove "medianage" and re-run the vif test 
vif_vals <- vif(lm(log(popunemplo +1) ~ log(1+popinlabou) +
                      log(1+onlylessth) + log(1+master) + 
                      log(1+africaninl) + log(1+asianinlab) + 
                      log(1+hispanicin), 
                    data = nycDatNoNA))

print(vif_vals)
```

```{r}
# Bandwidth 
bwVal <- GWmodel::bw.gwr(log(popunemplo +1) ~ log(1+popinlabou) +
                           log(1+onlylessth) + log(1+master) + 
                           log(1+africaninl) + log(1+asianinlab) + 
                           log(1+hispanicin),
                         data = nycDatNoNA %>% sf::as_Spatial(), 
                         approach = 'AICc', kernel = 'bisquare', 
                         adaptive = TRUE)
# note: # The code is using an “adaptive” neighborhood, which is defined as the k nearest neighbors based on Euclidean distances. 
# But then how do we know how many neighbors we should use for the local regression? Or what is best value for k? GWmodel::bw.gwr can help us search for such an optimal value based on the model fitness. 
```

## Run basic GWR 

```{r}
# Basic GWR
gwr.res <- gwr.basic(log(popunemplo +1) ~ log(1+popinlabou) +
                           log(1+onlylessth) + log(1+master) + 
                           log(1+africaninl) + log(1+asianinlab) + 
                           log(1+hispanicin),
                         data = nycDatNoNA %>% sf::as_Spatial(), 
                     bw = bwVal, kernel = "bisquare", adaptive = TRUE)
print(gwr.res)
```

While the pseudo-R square values improve a lot from GWR, the most valuable application of GWR outputs is to examine the spatial variations of the estimated model coefficients.

```{r}
# examine the spatial variations of the estimated model coefficients

# The gwr.res$SDF is a Spatial*DataFrame that contain all the coefficients estimated 
# at each sptial feature. We can simply map it out as a sp object
names(gwr.res$SDF)

# The SDF object in the gwr.basic resutls contains:
# + estimated coefficients
# + predicted Y calues
# + coefficient standard errors
# + t-values. 
# All these values are organized in a Spatial*DataFrame that can be directly mapped.
```

## Mapping the result

```{r}
# Because the spplot cannot handle those unconventional column names, we need to do extra works.
spGWRData <- gwr.res$SDF@data
spGWRData$coefMaster <- spGWRData$`log(1 + master)`

# Calculate the p value from (student) t value
spGWRData$pMaster <- 2*pt(-abs(spGWRData$`log(1 + master)_TV`), df = dim(spGWRData)[1] -1)

spGWRData$coefAfrican <- spGWRData$`log(1 + africaninl)`; 
spGWRData$coefHighschool <- spGWRData$`log(1 + onlylessth)`

spGWR <- gwr.res$SDF; 
spGWR@data <- spGWRData;
```

```{r}
spplot(spGWR, 'Local_R2', main="Local R Squared")
```

```{r}
# This show how the coefficients of the master degree on unemplyment change over space. 
spplot(spGWR, 'coefMaster', main="Estimated Coefficients of Master Degree on Unemployment")
```

```{r}
# But this variable is not signficiant everywhere.
spplot(spGWR, 'pMaster', main="p-value of Master Degree on Unemployment")
```

```{r}
# Another factor of highschool
spplot(spGWR, 'coefHighschool', main="Estimated Coefficients of Highshcool or Less on Unemployment")
```

