---
title: "Spatial Error Model"
output: html_document
---

# Spatial Error Model

The spatial error model handles the spatial autocorrelation in the residuals. The idea is that such errors (residuals from regression) are autocorrelated in that the error from one spatial feature can be modeled as a weighted average of the errors of its neighbors. In other words, such errors have spatial autocorrelation. 

## Dataset  

```{r}
nycDat <- sf::st_read('./data/nyctract_acs/NYC_Tract_ACS2008_12.shp')
```

```{r}
sf::st_crs(nycDat) <- 4326
```

## Weight matrix (Queen matrix as defaults)

```{r}
nycNbList <- nycDat %>% 
  spdep::poly2nb() %>%    # spdep::poly2nb produces neighboring relationships
  spdep::nb2listw(zero.policy = TRUE) # turns relationships into weights
```

```{r}
# view weight matrix (as a List) 
str(nycNbList)
View(nycNbList)

nycNbList$style
nycNbList$neighbours
nycNbList$weights
```

## OLS regression 

```{r}
# Prepare data: covert some factors to numeric values
nycDat <- nycDat %>% dplyr::mutate(medianage = medianage %>% as.character() %>% as.numeric(),
                          househol_1 = househol_1 %>% as.character() %>% as.numeric())
```

```{r}
# A simple linear regression on unemployed population

olsRslt <- lm(log(popunemplo +1) ~  # using a log-transformed approach to stabilize variance
                log(1+popinlabou) +
                log(1+onlylessth) + 
                log(1+master) + 
                log(1+africaninl) + 
                log(1+asianinlab) + 
                log(1+hispanicin) + 
                nycDat$medianage,
              data = nycDat)

summary(olsRslt)
```

```{r}
lmResiduals <- rep(0, length(nycDat$popunemplo)) # vector to store the residual
resIndex <- olsRslt$residuals %>% names() %>% as.integer() 
lmResiduals[resIndex] <- olsRslt$residuals
```

```{r}
# Test if there is spatial autocorrelation in the regression residuals (errors).
nycNbList %>%
  spdep::moran.test(lmResiduals, ., zero.policy = TRUE)
```

```{r}
# Choropleth Map of OLS Residuals

# Merge OLS residuals with the spatial dataset
nycDat$ols_residuals <- lmResiduals 

olsplot <- ggplot(nycDat) +
  geom_sf(aes(fill = ols_residuals), color = "white", size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(title = "OLS Residuals - Spatial Distribution",
       fill = "Residuals")
```

```{r}
ggsave("./plot/sem/ols_plot.png", plot = olsplot, width = 10, height = 6, dpi = 300)
```

# SEM 

```{r}
# Use spatialreg::errorsarlm to run the same model
serrRslt <- spatialreg::errorsarlm(log(popunemplo +1) ~ log(1+popinlabou) +
                log(1+onlylessth) + log(1+master) + 
                log(1+africaninl) + log(1+asianinlab) + 
                log(1+hispanicin) + medianage,
              data = nycDat,
              listw = nycNbList,
              zero.policy = TRUE, 
              na.action = na.omit);

summary(serrRslt)
```

```{r}
# Derive the residuals from the regression
seResiduals <- rep(0, length(nycDat$popunemplo))
resIndex <- serrRslt$residuals %>% names() %>% as.integer() 
seResiduals[resIndex] <- serrRslt$residuals
```

```{r}
# Test if there is spatial autocorrelation in the regression residuals (errors).
nycNbList %>%
  spdep::moran.test(seResiduals, ., zero.policy = TRUE) # seResiduals
```

```{r}
# Choropleth Map of SEM Residuals
nycDat$sem_residuals <- seResiduals 

seResiduals_plot <- ggplot(nycDat) +
  geom_sf(aes(fill = sem_residuals), color = "white", size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(title = "SEM Residuals - Spatial Distribution",
       fill = "Residuals")
```

```{r}
ggsave("./plot/sem/seResiduals_plot.png", plot = seResiduals_plot, width = 10, height = 6, dpi = 300)
```

# Just for fun with maps

```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(gridExtra)  # For arranging multiple plots

# Merge residuals with spatial dataset
nycDat$ols_residuals <- lmResiduals 
nycDat$sem_residuals <- seResiduals 

# Create OLS Residuals Map
ols_map <- ggplot(nycDat) +
  geom_sf(aes(fill = ols_residuals), color = "white", size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(title = "OLS Residuals - Spatial Distribution",
       fill = "Residuals")

# Create SEM Residuals Map
sem_map <- ggplot(nycDat) +
  geom_sf(aes(fill = sem_residuals), color = "white", size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(title = "SEM Residuals - Spatial Distribution",
       fill = "Residuals")

# Arrange maps side by side in a grid (2 columns)
grid.arrange(ols_map, sem_map, ncol = 2)
```

```{r}
# Adding North arrow and Scale bar 

# Check the CRS of your data
st_crs(nycDat)

# If your data is in degrees (EPSG:4326) and you want a more accurate scale bar, consider transforming it to a projected coordinate system like UTM before plotting
nycDat <- st_transform(nycDat, crs = 32618)  # Convert to UTM Zone 18N (for NYC)

# Add a scale bar with explicit unit control
ols_map <- ggplot(nycDat) +
  geom_sf(aes(fill = ols_residuals_cat), color = "white", size = 0.1) +
  scale_fill_manual(values = residual_colors) +
  theme_minimal() +
  labs(title = "OLS Residuals - Categorized", fill = "Residuals") +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  annotation_scale(location = "br", width_hint = 0.4, 
                   bar_cols = c("black", "white"), 
                   unit_category = "metric")  # Ensure metric units

# Print the map
ols_map
```

```{r}
# Function to categorize residuals
classify_residuals <- function(residuals) {
  cut(residuals, 
      breaks = c(-Inf, -2.3, -0.3, 0.0, 0.2, 0.4, Inf),  
      labels = c("Very Low (< -2.3)", 
                 "Low (-2.3 to -0.3)", 
                 "Neutral (-0.3 to 0.0)", 
                 "High (0.0 to 0.2)", 
                 "Very High (0.2 to 0.4)", 
                 "Extremely High (> 0.4)"),
      include.lowest = TRUE)
}

# Apply classification to residuals
nycDat <- nycDat %>%
  mutate(ols_residuals_cat = classify_residuals(lmResiduals),
         sem_residuals_cat = classify_residuals(seResiduals))

# Define color scale
residual_colors <- c("Very Low (< -2.3)" = "red", 
                     "Low (-2.3 to -0.3)" = "orange",
                     "Neutral (-0.3 to 0.0)" = "yellow", 
                     "High (0.0 to 0.2)" = "lightblue", 
                     "Very High (0.2 to 0.4)" = "blue",
                     "Extremely High (> 0.4)" = "darkblue")

# Create OLS Residuals Map (Categorized)
ols_map <- ggplot(nycDat) +
  geom_sf(aes(fill = ols_residuals_cat), color = "white", size = 0.1) +
  scale_fill_manual(values = residual_colors) +
  theme_minimal() +
  labs(title = "OLS Residuals - Categorized",
       fill = "Residuals")

# Create SEM Residuals Map (Categorized)
sem_map <- ggplot(nycDat) +
  geom_sf(aes(fill = sem_residuals_cat), color = "white", size = 0.1) +
  scale_fill_manual(values = residual_colors) +
  theme_minimal() +
  labs(title = "SEM Residuals - Categorized",
       fill = "Residuals")

# Arrange maps side by side in a grid (2 columns)
grid <- grid.arrange(ols_map, sem_map, ncol = 2)
```

```{r}
ggsave("./plot/sem/grid_ols_sem.png", plot = grid, width = 10, height = 6, dpi = 300)
```
