---
title: "2. Local Indicators of Spatial Autocorrelation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective 

The method is applied to New York City census tract data. For instance, the unemployment rate is analysed using LISA to identify which tracts are part of significant clusters (e.g. high unemployment surrounded by high unemployment) or outliers (e.g. low unemployment in a generally high unemployment area).

LISA can help identify clusters of high or low values as well as outliers that are surrounded by opposite values. As analysed with Global Indicators of Spatial Autocorrelation (look at Moran'I Plot), LISA could identify: 
- High-High (HH) clusters – areas with high values surrounded by high values (hotspots).
- Low-Low (LL) clusters – areas with low values surrounded by low values (coldspots).
- High-Low (HL) outliers – high values surrounded by low values.
- Low-High (LH) outliers – low values surrounded by high values.
- Insignificant areas – no significant spatial patterns.

## LISA  

```{r}
lisaRslt <- spdep::localmoran(nycDat$UNEMP_RATE, # localmoran 
                              nycNbList, 
                              zero.policy = TRUE, 
                              na.action = na.omit)
```

```{r}
dim(nycDat) 
dim(lisaRslt) 
```

```{r}
head(lisaRslt)
```

```{r}
significanceLevel <- 0.05 # 95% confidence

meanVal <- mean(nycDat$UNEMP_RATE)

lisaRslt %<>% tibble::as_tibble() %>%
  magrittr::set_colnames(c("Ii","E.Ii","Var.Ii","Z.Ii","Pr(z > 0)")) %>%
  dplyr::mutate(coType = dplyr::case_when(  # CoType - define Cluster/outliers types for each spatial feature in data, compare with mean value 
    `Pr(z > 0)` > 0.05 ~ "Insignificant",
    `Pr(z > 0)` <= 0.05 & Ii > 0 & nycDat$UNEMP_RATE >= meanVal ~ "HH",
    `Pr(z > 0)` <= 0.05 & Ii > 0 & nycDat$UNEMP_RATE < meanVal ~ "LL",
    `Pr(z > 0)` <= 0.05 & Ii < 0 & nycDat$UNEMP_RATE >= meanVal ~ "HL",
    `Pr(z > 0)` <= 0.05 & Ii < 0 & nycDat$UNEMP_RATE < meanVal ~ "LH"
  ))
```

```{r}
nycDat$coType <- lisaRslt$coType %>% tidyr::replace_na("Insignificant")
```

```{r}
p1 <- ggplot(nycDat) +
  geom_sf(aes(fill=coType),color = 'lightgrey') +
  scale_fill_manual(values = c('red','brown','NA','blue','cyan'), name='Clusters & \nOutliers') +
  labs(title = "LISA Clusters: Unemployment in NYC at Census Tract Level") +
  theme_minimal()
```

```{r}
ggsave("./plot/local indicator/local indicator_nyc.png", plot = p2, width = 10, height = 6, dpi = 300)
```

## Local clusters (hotspots, coldspots) & outliers

```{r}
# Define this as a function for 'plotCOType()"

plotCOType <- function(varName, titleText, cols=1:5) {
  varVals <- nycDat[[varName]] %>% as.character() %>% as.numeric()
  lisaRslt <- spdep::localmoran(varVals, nycNbList, 
                                zero.policy = TRUE, na.action = na.exclude)
  significanceLevel <- 0.05; # 95% confidence
  meanVal <- mean(varVals, na.rm=TRUE);
  
  lisaRslt %<>% tibble::as_tibble() %>%
    magrittr::set_colnames(c("Ii","E.Ii","Var.Ii","Z.Ii","Pr(z > 0)")) %>%
    dplyr::mutate(coType = dplyr::case_when(
      `Pr(z > 0)` > 0.05 ~ "Insignificant",
      `Pr(z > 0)` <= 0.05 & Ii >= 0 & varVals >= meanVal ~ "HH",
      `Pr(z > 0)` <= 0.05 & Ii >= 0 & varVals < meanVal ~ "LL",
      `Pr(z > 0)` <= 0.05 & Ii < 0 & varVals >= meanVal ~ "HL",
      `Pr(z > 0)` <= 0.05 & Ii < 0 & varVals < meanVal ~ "LH"
    ))
  
  # add this coType to original sf data
  nycDat$coType <- lisaRslt$coType %>% tidyr::replace_na("Insignificant")
  
  ggplot(nycDat) +
    geom_sf(aes(fill=coType),color = 'lightgrey') +
    scale_fill_manual(values = c('red','brown','NA','blue','cyan')[cols], name='Clusters & \nOutliers') +
    labs(title = titleText)
}
```

```{r}
p2 <- plotCOType('medianinco', "Median Household Income at Census Tract Level") # using varibale "medianinco"
```

```{r}
ggsave("./plot/local indicator/medianinco_nyc.png", plot = p2, width = 10, height = 6, dpi = 300)
```

```{r}
p3 <- plotCOType('asian', 'Asian Population at Census Tract Level', c(1,3,4,5)) 
```

```{r}
ggsave("./plot/local indicator/asian_nyc.png", plot = p3, width = 10, height = 6, dpi = 300)
```

```{r}
gridExtra::grid.arrange(p1, p2, ncol = 2) # Arrange All Plots in a Grid Layout for Comparison
```
